---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: /home/josh/School_16-17/Q2/info498c/project/global-gag-rule/paper/templates/svm-latex-ms.tex
title: "Global Effects Of The Mexico City Policy"
author:
- name: Joshua Keating, Peter Freschi, and John Harrison
  affiliation: University of Washington
abstract: "This document provides an introduction to R Markdown, argues for its benefits, and presents a sample manuscript template intended for an academic audience. I include basic syntax to R Markdown and a minimal working example of how the analysis itself can be conducted within R with the `knitr` package."
keywords: "Mexico City policy, Global Gag Rule"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
biblio-style: apsr
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

WHO_data <- read.csv("../data/sub-saharan-exposure-and-maternal-mortality", stringsAsFactors = FALSE)
colnames(WHO_data)[5] <- "Number_Abortions"
WHO.just.high <- WHO_data %>% filter(Exposure == "High")
WHO.just.low <- WHO_data %>% filter(Exposure == "Low")

```

# Introduction

The issue of abortion in the United States is one that is fiercely debated and divides large segments of the population. These divides are inherently partisan and therefore have led to legislation. In 1984 the Reagan administration signed into law the “Mexico City policy”, named for the city in which it was signed. The policy requires all non governmental organizations operating abroad to refrain from performing, advising on or endorsing abortion as a method of family planning if they wish to continue to receive receive federal funding. The policy has had a tumultuous history, either being reinstated or rescinded depending on the party affiliation of the current president. Starting with Bill Clinton in 1993 the policy was rescinded, then restored by George W. Bush in 2001, rescinded again by Barack Obama in 2009, and most recently restored by Donald J. Trump in 2017.

At it’s heart, the ban is propelled by the desire to limit the use of U.S. taxpayer dollars to pay for abortion or abortion-related services. The consequences of this ban include the termination of abortion related services such as; education, family planning, counseling, and training. The all or nothing nature of the ban would also cut off funds for non abortion related health services that would be offered at non governmental health providers. This presents some serious public health concerns for areas of the world where these services are essential for maternal health and prenatal care. 

We hypothesized that the institution of the ban could lead to a reduction in family planning services which could then lead to an increase in induced abortions. By restricting the flow of funds into health clinics that provide services related to abortion among others, the U.S. government restricts the amount of care that can be provided to those utilizing those services. The aim of this project was to gather, analyze, and present data on the past effects of the Global Gag Rule and create a resource that assists our audience in understanding the scope of the policy, its observable impact on women’s health, and the pathways through which it influences public health policy decisions outside of the United States.

# Related Work

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat


This is an example of a footer[^footerone] Here is another example[^footertwo] This is useful for both debugging and the publication stage, when the editorial team will ask for the raw `.tex` so that they could render it and later provide page proofs.

[^footerone]: Testing custom footer
[^footertwo]: testting second footer


# Methods

Heres some emmbeded code

```{r, tidy=TRUE}
# lmfit <- lm(Maternal_Mortality ~ Exposure + Num.Ab, data=full.exposure.mm, na.action = na.omit)
# lmfit
# preds <- predict(lmfit)
# plot(preds, full.exposure.mm$Maternal_Mortality[!is.na(full.exposure.mm$Num.Ab)])

tmp <- "test"

```

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat

need to fix this spacing here


```{r, echo = FALSE, warning = FALSE, fig.cap="Abortion Rates For Both Levels Of Exposure"}

Who.by.exp <- ggplot(NULL) +
    geom_smooth(data = WHO.just.high, method = "loess", se = FALSE, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_smooth(data = WHO.just.low, method = "loess", se = FALSE, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_point(data = WHO.just.high, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_point(data = WHO.just.low, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    labs(x = "Year", y = "Induced Abortion") +
    geom_vline(xintercept = 2001, linetype="dotted") + 
    theme_bw()

Who.by.exp

```


# Results

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat

# Discussion

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat

```{r, echo = FALSE, warning = FALSE, fig.cap="Test"}

Who.by.exp <- ggplot(NULL) +
    geom_smooth(data = WHO.just.high, method = "loess", se = FALSE, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_smooth(data = WHO.just.low, method = "loess", se = FALSE, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_point(data = WHO.just.high, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    geom_point(data = WHO.just.low, aes(x=Year, y=Number_Abortions, colour = Exposure)) + 
    labs(x = "Year", y = "Induced Abortion") +
    geom_vline(xintercept = 2001, linetype="dotted") + 
    theme_bw()

Who.by.exp

```


# Future Work

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat


<!--
# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent
-->
